---
title: "Q2"
output: html_document
date: "2025-03-30"
---



#Part 1: Neighbourhood-joining tree: 

Loading required packages:

```{r, echo=FALSE}
library(ggplot2)
library(ape)
library(ggtree)
library(vegan)
library(dplyr)
library(tidyr)
```

Loading in the dataset:

```{r}
ticks <- read.csv("432 Final Combined Data.csv")

Samples20 <- read.csv("samples20.csv")
```


processesing the dataset, filtering out low-abundance samples, and visualizing the distribution of remaining data:

```{r}
write.table(ticks, file = "OTUfile.txt", sep = "\t", row.names = F, col.names = T, quote = F)
OTUticks = read.delim("./OTUfile.txt",
                       header = T, sep = "\t",
                       row.names = "Sample")

OTUticks2 <- OTUticks %>%
  select(-c(Location))

OTUticks3 <- rowSums(OTUticks2) < 5
sum(OTUticks3)


OTUred <- OTUticks2[!OTUticks3,]

ggplot() + 
  geom_histogram(aes(x = rowSums(OTUred)))
```




Generating the OTU dissimilarity matrix based on Bray-Curtis dissimilarity:

```{r}
OTUred[OTUred > 0] <- 1

OTUTickdistance2 <- vegdist(OTUred, method = "bray", binary = F)

OTUtree2 <- nj(OTUTickdistance2)

```


Creating the neighbourhood-joining tree: 

```{r}
ggtree(OTUtree2, layout = "circular") %<+% Samples20 + 
  geom_tiplab(aes(colour = Location), size = 2) + 
  theme(legend.position = "right")  + 
  theme_void() +
  xlim(-0.75,0.75) + 
  scale_color_discrete(name = "Sampling Location",
                       labels = c("Control", "Lemoine Point", "Murphy's Point", "QUBS"))
```


#PART 2: NMDS plot:

Creating NMDS matrix: 

```{r, echo=FALSE}
set.seed(13)

OTUtickdistance <- dist(OTUred, method = 'binary')

TickNMDS <- metaMDS(OTUtickdistance, k = 2)
```


Generating NMDS plot: 

```{r}
PTickDat <- data.frame(NMDSA1 = TickNMDS$points[,1],
                   NMDSA2 = TickNMDS$points[,2],
                   Sample = row.names(OTUred))

PTickDat <- merge(PTickDat, ticks, by = "Sample", all.x = T, all.y = F)

ggplot(aes(x = NMDSA1, y = NMDSA2, colour = Location),
       alpha = I(0.6), data = PTickDat) + 
  geom_point(size = 3) + 
  theme_bw() + 
  theme(axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14)) + 
  theme(legend.position = "bottom") + 
  scale_color_discrete(name = "Sampling Location",
                       labels = c("Control", "Lemoine Point", "Murphy's Point", "QUBS"))
```


#PART 3: Statistical Analysis- PERMANOVA

Creating the proper data set for the PERMANOVA. Log-transforming to account for many zeros in dataset:

```{r}
asv_matrix <- ticks[, -(1:2)]
asv_matrix <- decostand(asv_matrix, method = "hellinger")
asv_dist <- vegdist(asv_matrix, method = "bray")
asv_matrix <- log1p(asv_matrix)

```


Performing the PERMANOVA:

```{r}
permanova_results <- adonis2(asv_dist ~ Location, data = ticks, permutations = 999)
print(permanova_results)

```




