
## Question 3

```{r message=F, warning=F}
library(dplyr)
library(ggplot2)
library(stringr)
library(cowplot)
library(ggrepel)
source("http://bit.ly/theme_pub")
```

```{r}
data = read.table("./data/rawGeneExpression.txt", header=T, sep = "\t")

# remove unwanted columns & rename
data = data %>% 
  select(1:5) %>%
  rename(Bacteria = X) %>%
  mutate(across(2:5, as.numeric))

data$Bacteria = str_extract(data$Bacteria, "^[^;]+") # removing extra stuff from bacteria name
```

```{r warning=F}
PCA_data = as.data.frame(scale(data[2:5])) # scale data

PCA = princomp(PCA_data, cor=T) # run PCA

PCA_data$PC1 = PCA$scores[,1] # getting PC axes for plot
PCA_data$PC2 = PCA$scores[,2]
PCA_data$Bacteria = data$Bacteria # add names back for colour coding

ggplot(data=PCA_data, aes(x=PC1, y=PC2, colour=Bacteria)) + geom_point() + theme_pub() # plot is bad, we will remove outliers
```

```{r}
summary(PCA)
PCA$loadings
```


```{r warning=FALSE}
PCA_data = PCA_data[-which(PCA_data$PC1 > -0.15),] # remove outliers, only 41 observation difference. We are not intersted in the far away ones anyways. 

# plot
x = ggplot(data=PCA_data, aes(x=PC1, y=PC2, colour=Bacteria)) + geom_point() + theme(legend.text = element_text(size = 8), legend.key.size = unit(0.5, "cm"))

# get and print legend (legend too big to print with graph)
legend = get_legend(x) 
plot_grid(legend)
```

```{r fig.cap = "PCA analysis of bacterial species found in four different tick species near Kingston, Ontario."}
# get labels
labels = function(x) {
  words = unlist(strsplit(x, " "))
  abbr = paste0(substr(words, 1, 1), collapse = ".")
  return(abbr)
}

# add labels
PCA_data = PCA_data %>%
    mutate(label = sapply(Bacteria, labels))

# plot with labels
options(ggrepel.max.overlaps = Inf) # need this for all labels

x + theme_pub() + geom_text_repel(data = PCA_data, aes(label = ifelse(Bacteria == "Rickettsia buchneri", "", label)), size = 3, alpha = 0.5,)

# the unlabeled purple points are Rickettsia buchneri
```

PCA analysis of bacterial species found in four different tick species. Each point represents a bacterial species, with proximity to others indicating co-occurrence across similar tick samples. 

However, there are only four tick samples. This causes all bacteria found to cluster relatively closely. To get more precise idea, we can use correlations between the three bacteria of interest and the other bacteria found in the tick samples. 

```{r warning=F}
# set up data for correlation calculations
grouped_data = PCA_data %>%
  group_by(Bacteria) %>%
  summarise(across(starts_with("X"), mean))

rownames(grouped_data) = grouped_data$Bacteria
cor_data = t(grouped_data)
cor_data = as.data.frame(cor_data[-1,])
cor_data[] = lapply(cor_data, as.numeric)
```

```{r}
# correlation with Anaplasma phagocytophilum

correlations_ap = data.frame(Comparison = character(), Correlation = numeric())

for (i in 1:ncol(cor_data)){
  x = cor(cor_data$`Anaplasma phagocytophilum`, cor_data[[i]])
  y = data.frame(Comparison = sprintf("Anaplasma phagocytophilum vs. %s", colnames(cor_data[i])), Correlation = x)
  correlations_ap = rbind(correlations_ap, y)
}

correlations_ap = correlations_ap %>%
  arrange(desc(Correlation)) %>%
  slice(-1) # the first row is the bacteria vs itself
```

```{r}
# correlation with Borreliella burgdorferi

correlations_bb = data.frame(Comparison = character(), Correlation = numeric())

for (i in 1:ncol(cor_data)){
  x = cor(cor_data$`Borreliella burgdorferi`, cor_data[[i]])
  y = data.frame(Comparison = sprintf("Borreliella burgorferi vs. %s", colnames(cor_data[i])), Correlation = x)
  correlations_bb = rbind(correlations_bb, y)
}

correlations_bb = correlations_bb %>%
  arrange(desc(Correlation)) %>%
  slice(-1) # the first row is the bacteria vs itself
```

```{r}
# correlation with Borrelia miyamotoi

correlations_bm = data.frame(Comparison = character(), Correlation = numeric())

for (i in 1:ncol(cor_data)){
  x = cor(cor_data$`Borrelia miyamotoi`, cor_data[[i]])
  y = data.frame(Comparison = sprintf("Borrelia miyamotoi vs. %s", colnames(cor_data[i])), Correlation = x)
  correlations_bm = rbind(correlations_bm, y)
}

correlations_bm = correlations_bm %>%
  arrange(desc(Correlation)) %>%
  slice(-1) # the first row is the bacteria vs itself
```


```{r}
# make table for poster
newdata = data.frame()
newdata = data.frame(Comparison = correlations_ap$Comparison[1:3], Correlation = correlations_ap$Correlation[1:3]) %>%
  rbind(newdata, data.frame(Comparison = correlations_bb$Comparison[1:3], Correlation = correlations_bb$Correlation[1:3])) %>%
  rbind(newdata, data.frame(Comparison = correlations_bm$Comparison[1:3], Correlation = correlations_bm$Correlation[1:3]))

knitr::kable(newdata)
```

All correlations are high due to there being only four tick samples. Nearly all are within 0.1 of being perfectly correlated. By another method:

45F is the only tick to significantly express Borreliella burgorferi, the primary diagnostic pathogen for LD. 38F also does, but only 1 instance compared to 13. We will find all bacteria that were only expressed in tick 45F.

```{r}
bac_45F = list()

for (i in 1:nrow(data)){
  if (sum(data[i,2:4]) == 0 &
      data[i,5] != 0  &
      !(data$Bacteria[i] %in% c("Anaplasma phagocytophilum",
                                  "Borreliella burgdorferi",
                                  "Rickettsia buchneri",
                                  "Borrelia miyamotoi"))){
    bac_45F[[length(bac_45F)+1]] = data$Bacteria[i]
  }
}

bac_45F
```

The displayed bacteria are the ones only found in tick 45F, which is the only tick expressing Borreliella burgorferi significantly. Therefore these bacteria could be potential bio-markers for LD as well. Further analysis with more ticks could verify these conclusions. 








